A radiocarbon data based evaluation of Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe

### 2019-01-10 article/01_Introduction.Rmd
better integration of evolutionary archaeology

Archaeology has a unique perspective on *Cultural evolution* [@garveyCurrentPotentialRoles2018]: It is concerned with human behaviour over big areas and long timescales. It also works with solid proxies of cultural expression: material remains -- individual lost or discarded artefacts, the ruins of past architecture and all more inconspicuous by-products of living. We rightfully expect humans to learn handcraft as well as building techniques and traditions  -- both functionally meaningful and seemingly pointless -- from others close to them. The similarity of historically and spatially close material remains can, therefore, be explained as a direct result of social interaction [@lymanChronometersUnitsEarly2000; @lymanCultureHistoricalBiological2001; @lymanRiseFallCulture1997; @obrienEpistemologicalNatureArchaeological2002]. Questions of *Cultural Transmission* form the core of archaeology.

In this context a distinction of culture in function and style became an essential question for *Evolutionary archaeology*, explicitly elaborated already by @dunnell1978style. Some types of human behaviour increase well-being and simplify survival, like specific hunting techniques, superior tools or sowing more fruitful crops. Without resorting to the inappropriate simplifications of Evolutionism, their diffusion can be expected to increase over time. On the other hand, some behaviours are not related to survival and reproductive capacity, like e.g. ceramic decorations, traditional costume fashion or jewellery design. Despite (or because?) their seemingly secondary importance these characteristics often pass through a breathtakingly fast development and serve archaeologists as a proxy for relative chronology. Dunnell postulates that the distribution and emergence of these stylistic cultural traits behave randomly on a population scale level. The stochastic processes that are at work are an important object of investigation and discussed as *Neutral Transmission* or *Stylistic variability* [@bentleyRandomDriftCulture2004; @lipoNeutralityStyleDrift2001; @neimanStylisticVariationEvolutionary1995; @porcicExploringEffectsAssemblage2015; @rogersNaturalSelectionCultural2008; @shennanCeramicStyleChange2001].

@neimanStylisticVariationEvolutionary1995 takes a closer look at this processes and develops a simple framework for the simulation of non-functional idea propagation. Neiman identifies the essential phenomena *drift* (random dominance of similar traits) and *flow* (intergroup transmission) in population genetics to be also relevant for cultural traits in archaeology. Furthermore, he implements a measure of cultural distance to apply his theoretical insights to real-world data. The workflow suggested by Neiman can be partially applied to the burial rites data examined in this article and was an important methodological inspiration: As burial traditions are ritualised behaviour and most of the time heavily linked to religious and spiritual beliefs they can be assumed to be stylistic traits as defined by @dunnell1978style. Besides @neimanStylisticVariationEvolutionary1995, another vital template for the comparison of cultural and spatial was provided by @porcicSimulatingCulturalTransmission2014.

>>>>

Archaeology has a unique perspective on Cultural evolution [@garveyCurrentPotentialRoles2018]: It is concerned with human behaviour over big areas and long timescales. It also works with solid proxies of cultural expression: material remains -- individual lost or discarded artefacts, the ruins of past architecture and all more inconspicuous by-products of living. We rightfully expect humans to learn handcraft as well as building techniques and traditions  -- both functionally meaningful and seemingly pointless -- from others close to them. The similarity of historically and spatially close material remains can, therefore, be explained as a direct result of social interaction [@lymanChronometersUnitsEarly2000; @lymanCultureHistoricalBiological2001; @lymanRiseFallCulture1997; @obrienEpistemologicalNatureArchaeological2002]. Questions of *Cultural Transmission* form the core of archaeology.

*Evolutionary archaeology*, the branch of archaeology that explicitly adopted the terminology of Cultural evolution theory, established an important distinction of cultural behaviour in *function* and *style* (@dunnell1978style). Some types of human behaviour increase well-being and simplify survival, like specific hunting techniques, superior tools or sowing more fruitful crops. Without resorting to the inappropriate simplifications of Evolutionism, their diffusion can be expected to increase over time if we assume reasonable human actors. On the other hand, some behaviours are not related to survival and reproductive capacity, like e.g. ceramic decorations, traditional costume fashion or jewellery design. Despite or because their seemingly subordinate importance, these characteristics often pass through a breathtakingly fast development. They often serve archaeologists as a proxy for precise relative chronology. Dunnell postulates that the distribution and emergence of these stylistic cultural traits behave randomly on a population scale level. The stochastic processes that are at work are an important object of investigation and discussed as *Neutral Transmission* or *Stylistic variability* [@bentleyRandomDriftCulture2004; @lipoNeutralityStyleDrift2001; @neimanStylisticVariationEvolutionary1995; @porcicExploringEffectsAssemblage2015; @rogersNaturalSelectionCultural2008; @shennanCeramicStyleChange2001]. @neimanStylisticVariationEvolutionary1995 takes a closer look at this processes and develops a simple framework for the simulation of non-functional idea propagation. He identifies the essential phenomena *drift* (random dominance of similar traits) and *flow* (intergroup transmission) in population genetics to be also relevant for cultural traits in archaeology. Furthermore, he implements a measure of cultural distance to apply his theoretical insights to real-world data.

The workflow suggested by Neiman can be partially applied to the burial rites data examined in this article and was an important methodological inspiration: As burial traditions are ritualised behaviour and most of the time heavily linked to religious and spiritual beliefs they can be assumed to be stylistic traits as defined by @dunnell1978style. Besides @neimanStylisticVariationEvolutionary1995, another vital template for the comparison of cultural and spatial distance was provided by @porcicSimulatingCulturalTransmission2014.
